Introducing ICBe: Very High Recall and Precision Event Extraction from Narratives about International Crises

### 2022-09-30 replication_paper/political_analysis_draft/ICBe_political_analysis_submission_rmd.Rmd
trim down

nocite: |
  @fazalStateDeathPolitics2011, @ryanNationStatesEmpiresWars2021, @spruytSovereignStateIts1996, @haffarEmergentPeacemakersCataloguing2002, @kuznetsovTheoryPracticeParadiplomacy2014,  @bushDensityDeclineFounding2019, @mcclearyPrivateVoluntaryOrganizations2008, @olterIntergovernmentalOrganizationsIGOs2021, @benyehudaEthnicActorsInternational2006, @buenodemesquitaDomesticExplanationsInternational2012, @lacinaExplainingSeverityCivil2006, @lacinaMonitoringTrendsGlobal2005, @carafanoMeasuringMilitaryPower2014, @goertzMeasuringMilitaryAllocations1986, @mcnabbcochranMeasuringMilitaryEffectiveness2017, @gartzkeCrossDomainDeterrenceStrategy2019, @lanoszkaLandpowerAmericanCredibility2016, @lindsayPoliticsManyOther2020, @carterStrategyTerritorialConflict2010, @luptonReexaminingReputationResolve2018, @goldgeierPsychologyInternationalRelations2001, @mercerProspectTheoryPolitical2005, @steinEvaluatingWarOutcomes1980, @sullivanWarAimsWar2007, @ramsayInformationUncertaintyWar2017, @yarhi-miloEyeBeholderHow2013, @oneillInternationalNegotiationConceptual2018, @tragerDiplomacyWarPeace2016, @zartmanEscalationNegotiationInternational2005, @davisThreatsPromisesPursuit2000, @sechserMilitarizedCompellentThreats2011, @powellBargainingTheoryInternational2002, @goemansIntroducingArchigosDataset2009, @harknessEthnicArmyState2016, @jonesHitMissEffect2009, @matanockElectingPeaceCivil2017, @powellGlobalInstancesCoups2011, @chenowethIntroducingNonviolentAction2019, @eckOneSidedViolenceCivilians2007, @lafreeIntroducingGlobalTerrorism2007, @beardsleyMediationDilemma2011, @giblerMeasuringAlliancesCorrelates2004, @owsiakInternationalBorderAgreements2018, @leedsAllianceReliabilityTimes2003, @leedsDomesticPoliticalInstitutions1999, @demesquitaForeignAidPolicy2007, @yarhi-miloArmAllyPatron2016, @laiEffectsDifferentTypes2004, @allenUSGlobalMilitary2021, @fortnaPeaceTime2018, @minInterstateWarBattle2021, @levyCausesWar2011, @reiterHowWarsEnd2009, @frederickIssueCorrelatesWar2017, @henselChartingCourseConflict1996

@@

If we could record every important interaction between countries in all of diplomacy, military conflict, and international political economy, how much unique information would this chronicle amount to, and how surprised would we be to see something new? In other words, what is the entropy of international relations? This record could in principle be unbounded, but the central conceit of social science is that there are structural regularities that limit what actors can do, their best options, and even which actors are likely to survive [@brecherInternationalStudiesTwentieth1999; @reiterShouldWeLeave2015]. If so, then these events can be systematically measured, and accordingly, massive effort is expended in social science attempting to record these regularities.^[See work on crises [@brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020], militarized disputes [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018; @maozDyadicMilitarizedInterstate2019], wars [@sarkeesResortWar181620072010; @reiterRevisedLookInterstate2016], organized violence [@ralphsundbergUCDPGEDCodebook2016; @petterssonOrganizedViolence19892018], political violence [@raleighIntroducingACLEDArmed2010], sanctions [@felbermayrGlobalSanctionsData2020], trade [@barariDemocracyTradePolicy], and international agreements [@kinneDefenseCooperationAgreement2020; @owsiakInternationalBorderAgreements2018; @vabulasCooperationAutonomyBuilding2021], dispute resolution [@vabulasCooperationAutonomyBuilding2021; @frederickIssueCorrelatesWar2017], and diplomacy [@moyerWhatAreDrivers2020; @sechserMilitarizedCompellentThreats2011].] Thanks to improvements in natural language processing, more open-ended efforts have begun to capture entire unstructured streams of events from international news reports.^[See @liComprehensiveSurveySchemabased2021; @haltermanExtractingPoliticalEvents2020; @brandtPhoenixRealTimeEvent2018; @boscheeICEWSCodedEvent2015; @hegreIntroducingUCDPCandidate2020; @grantOUEventData2017. On event extraction from images and social media see @zhangCASMDeepLearningApproach2019 and @steinert-threlkeldFutureEventData2019.] How close these efforts are to accurately measuring all or even most of what is essential in international relations is an open empirical question, one for which we provide new evidence here.

Our contribution is a high coverage ontology and event dataset for key historical episodes in 20th and 21st-century international relations. We develop a large, flexible ontology of international events with the help of both human coders and natural language processing. We apply it sentence-by-sentence to an unusually high-quality corpus of historical narratives of international crises [@brecherInternationalStudiesTwentieth1999; @brecherCrisisEscalationWar2000; @wilkenfeldInterstateCrisesViolence2000; @jamesWhatWeKnow2019; @iakhnisCrisesWorldPolitics2019]. The result is a new lower bound estimate of how much actually happens between states during pivotal historical episodes. We then develop several methods for objectively gauging how well these event codings reconstruct the information contained in the original narrative. We conclude by benchmarking our event codings against several current state of the art event data collection efforts. We find that existing systems produce sequences of events that do not contain enough information to reconstruct the underlying historical episode. The underlying fine-grained variation in international affairs is unrecognizable through the lens of current quantification efforts.

@@

We consider the measurement task of absstracting discrete events about a historical episode in international relations. The easiest way to convey the task is with an example. Figure 1 shows a narrative account of the Cuban Missile Crisis (1962) alongside a mapping from each natural language sentence to discrete machine-readable abstractive events. Formally, a historical episode, $H$, is demarcated by a period of time $[T_{start}, T_{end}] \in T$, a set of Players $p \in P$, and a set of behaviors they undertook during that time $b \in B$. International Relations, $IR$, is the system of regularities that govern the strategic interactions that world actors make during a historical episode, given their available options, preferences, beliefs, and expectations of choices made by others. We observe neither $H$ nor $IR$ directly. Rather the Historical Record, $HR$, produces documents $d \in D$ containing some relevant and true (as well as irrelevant and untrue) information about behaviors that were undertaken recorded in the form of unstructured natural language text. The task is to combine informative priors about $IR$ with an unstructured corpus $D$ to produce a series of structured discrete events, $e \in E$, that have high coverage, precision, and recall over what actually took place in history, $H$.

@@

For our corpus, $D$, we select a set of unusually high-quality historical narratives from the International Crisis Behavior (ICB) project ($n=471$)(Supplementary Information (SI) Appendix 1.1)[@brecherInternationalCrisisBehavior2017; @brecher_study_1997]. Their domain is 20th and 21st-century crises, defined as a change in the type, or an increase in the intensity, of disruptive interaction with a heightened probability of military hostilities that destabilizes states' relationships or challenges the structure of the international system [@brecherCrisesWorldPolitics1982].^[On near crises see @iakhnisCrisesWorldPolitics2019.] Crises are a significant focus of detailed single case studies and case comparisons because they provide an opportunity to examine behaviors in IR short of, or at least prior to, full conflict  [@holsti1914Case1965; @paigeKoreanDecisionJune1968; @allisonEssenceDecisionExplaining1971; @snyderConflictNationsBargaining1977; @gavinHistorySecurityStudies2014; @georgeDeterrenceAmericanForeign1974; @brecherCrisesWorldPolitics1982; @gaddisExpandingDataBase1987; @brecherPatternsCrisisManagement1988]. Case selection was exhaustive based on a survey of world news archives and region experts, cross-checked against other databases of war and conflict, and non-English sources [@kangUSBiasStudy2019; @brecherInternationalCrisisBehavior2017 p. 59]. Each narrative was written by consensus by a small number of scholars, using a uniform coding scheme, with similar specificity [@hewittEngagingInternationalData2001]. The corpus is unique in IR because it is designed to be used in a downstream quantitative coding project.

@@

The second axis (columns) compares the very high ontological coverage of ICBe to existing state of the art systems in production and with global coverage. They begin with our contribution ICBe, alongside other event-level datasets including CAMEO dictionary lookup-based systems (Historical Phoenix [@althausClineCenterHistorical2019]; ICEWS [@boscheeICEWSCodedEvent2015; @hegreIntroducingUCDPCandidate2020]; Terrier [@grantOUEventData2017]), the Militarized Interstate Disputes Incidents dataset, and the UCDP-GED dataset [@ralphsundbergUCDPGEDCodebook2016; @petterssonOrganizedViolence19892018; @sundbergIntroducingUCDPGeoreferenced2013].^[Other related datasets that insufficiently overlap ICBe's domain for comparison include BCOW [@lengMilitarizedInterstateCrises1988], WEIS [@mcclellandWorldEventInteraction1978], CREON [@hermannComparativeResearchEvents1984], CASCON [@bloomfieldCASCONIIIComputeraided1989], SHERFACS [@shermanSHERFACSCrossParadigmHierarchical2000], Real-Time Phoenix [@brandtPhoenixRealTimeEvent2018], and COfEE [@balaliCOfEEComprehensiveOntology2021] (see histories in @merrittMeasuringEventsInternational1994 and @schrodtTwentyYearsKansas2006).] The final set of columns compares episode-level datasets beginning with the original ICB project [@brecherInternationalCrisisBehavior; @brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020]; the Militarized Interstate Disputes dataset [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018;@braithwaiteMIDLOCIntroducingMilitarized2010;@braithwaiteCodebookMilitarizedInterstate2009], and the Correlates of War [@sarkeesResortWar181620072010]. With the exception of large-scale CAMEO dictionary-based systems, the existing state of the art quantitative datasets ignore the vast majority of the information content found in international relations.^[See @balaliCOfEEComprehensiveOntology2021 for a recent review of ontological depth and availability of Gold Standard example text.]

@@

lit_review_alt <- lit_review_pull %>% 
                  mutate(Literature = paste0("",Literature %>% stringr::str_replace("; {0,}$","") ,"")) %>%
                  mutate(Literature = paste0("",Literature %>% stringr::str_replace(";",",") ,"")) %>%
                  #mutate(Literature = paste0("  [",Literature %>% stringr::str_replace("; {0,}$","") ,"]")) %>%
                  #mutate(Concept= paste0("**", Concept, "**", Literature)) %>%
                  mutate(Concept= paste0(Concept)) %>%
                  mutate(Literature = Literature %>% stringr::str_replace_all(stringr::fixed("NA"),"") ) %>%
                  mutate(Concept = Concept %>% stringr::str_replace(stringr::fixed(" ["),stringr::fixed(" - ["))) %>%
                  dplyr::select(-Codebook, -`Cameo Codes`, -`ICB Codes`, Literature, -`ICB Corpus`, -`Removed Literature`) 
                  #dplyr::filter(is.na(group))

@@

We release several open-source products along with supporting code and documentation to further advance the study of IR, event extraction, and natural language processing. The first is the International Crisis Behavior Events (ICBe) dataset, an event-level aggregation of what took place during the crises identified by the ICB project. These data are appropriate for statistical analysis of hard questions about the sequencing of events (e.g. escalation and de-escalation of conflicts).^[Using ICBe data @gannonOneIfLand2022 finds that cross-domain crises are shorter and less violent than same-domain crises.] Second, we provide a coder-level disaggregation with multiple codings of each sentence by experts and undergrads that allows for the introduction of uncertainty and human interpretation of events. Further, we release a direct mapping from the codings to the source text at the sentence level as a new resource for natural language processing. Finally, we provide a companion website that incorporates detailed visualizations of all of the data introduced here (www.crisisevents.org).

>>>>

nocite: |
  @fazalStateDeathPolitics2011, @spruytSovereignStateIts1996, @haffarEmergentPeacemakersCataloguing2002, @bushDensityDeclineFounding2019, @benyehudaEthnicActorsInternational2006, @lacinaExplainingSeverityCivil2006; @mcnabbcochranMeasuringMilitaryEffectiveness2017, @carafanoMeasuringMilitaryPower2014, @goertzMeasuringMilitaryAllocations1986, @gartzkeCrossDomainDeterrenceStrategy2019, @lindsayPoliticsManyOther2020, @carterStrategyTerritorialConflict2010, @goldgeierPsychologyInternationalRelations2001, @steinEvaluatingWarOutcomes1980, @sullivanWarAimsWar2007, @ramsayInformationUncertaintyWar2017, @yarhi-miloEyeBeholderHow2013, @oneillInternationalNegotiationConceptual2018, @zartmanEscalationNegotiationInternational2005, @sechserMilitarizedCompellentThreats2011, @powellBargainingTheoryInternational2002, @goemansIntroducingArchigosDataset2009, @powellGlobalInstancesCoups2011, @chenowethIntroducingNonviolentAction2019, @eckOneSidedViolenceCivilians2007, @lafreeIntroducingGlobalTerrorism2007, @beardsleyMediationDilemma2011, @giblerMeasuringAlliancesCorrelates2004, @owsiakInternationalBorderAgreements2018, @leedsAllianceReliabilityTimes2003, @leedsDomesticPoliticalInstitutions1999, @yarhi-miloArmAllyPatron2016, @laiEffectsDifferentTypes2004, @allenUSGlobalMilitary2021, @fortnaPeaceTime2018, @minInterstateWarBattle2021, @frederickIssueCorrelatesWar2017, @sarkeesResortWar181620072010, @reiterRevisedLookInterstate2016

@@

If we could record every important interaction between countries in all of diplomacy, military conflict, and international political economy, how much unique information would this chronicle amount to, and how surprised would we be to see something new? In other words, what is the entropy of international relations? This record could in principle be unbounded, but the central conceit of social science is that there are structural regularities that limit what actors can do, their best options, and even which actors are likely to survive [@brecherInternationalStudiesTwentieth1999; @reiterShouldWeLeave2015]. If so, then these events can be systematically measured, and accordingly, massive effort is expended in social science attempting to record these regularities.^[See work on crises [@brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020], militarized disputes [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018], wars [@sarkeesResortWar181620072010; @reiterRevisedLookInterstate2016], organized violence [@ralphsundbergUCDPGEDCodebook2016; @petterssonOrganizedViolence19892018], political violence [@raleighIntroducingACLEDArmed2010], sanctions [@felbermayrGlobalSanctionsData2020], and international agreements [@kinneDefenseCooperationAgreement2020; @owsiakInternationalBorderAgreements2018], dispute resolution [@frederickIssueCorrelatesWar2017], and diplomacy [@moyerWhatAreDrivers2020; @sechserMilitarizedCompellentThreats2011].] Thanks to improvements in natural language processing, more open-ended efforts have begun to capture entire unstructured streams of events from international news reports.^[See @liComprehensiveSurveySchemabased2021; @beielerGeneratingPoliticalEvent2016; @brandtPhoenixRealTimeEvent2018; @boscheeICEWSCodedEvent2015. On event extraction from images and social media see @zhangCASMDeepLearningApproach2019 and @steinert-threlkeldFutureEventData2019.] How close these efforts are to accurately measuring all or even most of what is essential in international relations is an open empirical question, one for which we provide new evidence here.

<!-- removedcites trade @barariDemocracyTradePolicy,  @vabulasCooperationAutonomyBuilding2021 -->
<!-- removedcites @hegreIntroducingUCDPCandidate2020; @grantOUEventData2017; @haltermanExtractingPoliticalEvents2020; @maozDyadicMilitarizedInterstate2019, @vabulasCooperationAutonomyBuilding2021;  -->

Our contribution is a high coverage ontology and event dataset for key historical episodes in 20th and 21st-century international relations. We develop a large, flexible ontology of international events with the help of both human coders and natural language processing. We apply it sentence-by-sentence to an unusually high-quality corpus of historical narratives of international crises [@brecherInternationalStudiesTwentieth1999; @brecherEscalationWarTwentieth2000; @wilkenfeldInterstateCrisesViolence2000; @brecherInternationalCrisisBehavior2021]. The result is a new lower bound estimate of how much actually happens between states during pivotal historical episodes. We then develop several methods for objectively gauging how well these event codings reconstruct the information contained in the original narrative. We conclude by benchmarking our event codings against several current state of the art event data collection efforts. We find that existing systems produce sequences of events that do not contain enough information to reconstruct the underlying historical episode. The underlying fine-grained variation in international affairs is unrecognizable through the lens of current quantification efforts.

<!-- removedcites  @jamesWhatWeKnow2019; @iakhnisCrisesWorldPolitics2019 -->

@@

We consider the measurement task of abstracting discrete events about a historical episode in international relations. The easiest way to convey the task is with an example. Figure 1 shows a narrative account of the Cuban Missile Crisis (1962) alongside a mapping from each natural language sentence to discrete machine-readable abstractive events. Formally, a historical episode, $H$, is demarcated by a period of time $[T_{start}, T_{end}] \in T$, a set of Players $p \in P$, and a set of behaviors they undertook during that time $b \in B$. International Relations, $IR$, is the system of regularities that govern the strategic interactions that world actors make during a historical episode, given their available options, preferences, beliefs, and expectations of choices made by others. We observe neither $H$ nor $IR$ directly. Rather the Historical Record, $HR$, produces documents $d \in D$ containing some relevant and true (as well as irrelevant and untrue) information about behaviors that were undertaken recorded in the form of unstructured natural language text. The task is to combine informative priors about $IR$ with an unstructured corpus $D$ to produce a series of structured discrete events, $e \in E$, that have high coverage, precision, and recall over what actually took place in history, $H$.

@@

For our corpus, $D$, we select a set of unusually high-quality historical narratives from the International Crisis Behavior (ICB) project ($n=471$)(Supplementary Information (SI) Appendix 1.1)[@brecherInternationalCrisisBehavior2021; @brecher_study_1997]. Their domain is 20th and 21st-century crises, defined as a change in the type, or an increase in the intensity, of disruptive interaction with a heightened probability of military hostilities that destabilizes states' relationships or challenges the structure of the international system [@brecherCrisesWorldPolitics1982]. Crises are a significant focus of detailed single case studies and case comparisons because they provide an opportunity to examine behaviors in IR short of, or at least prior to, full conflict  [@holsti1914Case1965; @paigeKoreanDecisionJune1968; @allisonEssenceDecisionExplaining1971; @gavinHistorySecurityStudies2014; @brecherCrisesWorldPolitics1982; @iakhnisCrisesWorldPolitics2019]. Case selection was exhaustive based on a survey of world news archives and region experts, cross-checked against other databases of war and conflict, and non-English sources [@kangUSBiasStudy2019; @brecherInternationalCrisisBehavior2021 p. 59]. Each narrative was written by consensus by a small number of scholars, using a uniform coding scheme, with similar specificity [@hewittEngagingInternationalData2001]. The corpus is unique in IR because it is designed to be used in a downstream quantitative coding project.

<!-- removedcites @snyderConflictNationsBargaining1977; @gaddisExpandingDataBase1987; @georgeDeterrenceAmericanForeign1974; @brecherPatternsCrisisManagement1988; -->

@@

The second axis (columns) compares the very high ontological coverage of ICBe to existing state of the art systems in production and with global coverage. They begin with our contribution ICBe, alongside other event-level datasets including CAMEO dictionary lookup-based systems (Historical Phoenix [@althausClineCenterHistorical2019]; ICEWS [@boscheeICEWSCodedEvent2015; @hegreIntroducingUCDPCandidate2020]; Terrier [@grantOUEventData2017]), the Militarized Interstate Disputes Incidents dataset, and the UCDP-GED dataset [@ralphsundbergUCDPGEDCodebook2016; @petterssonOrganizedViolence19892018; @sundbergIntroducingUCDPGeoreferenced2013].^[Other related datasets that insufficiently overlap ICBe's domain for comparison include BCOW [@lengMilitarizedInterstateCrises1988], WEIS [@mcclellandWorldEventInteraction1978], CREON [@hermannComparativeResearchEvents1984], CASCON [@bloomfieldCASCONIIIComputeraided1989], SHERFACS [@shermanSHERFACSCrossParadigmHierarchical2000], Real-Time Phoenix [@brandtPhoenixRealTimeEvent2018], and COfEE [@balaliCOfEEComprehensiveOntology2021] (see histories in @merrittMeasuringEventsInternational1994 and @schrodtTwentyYearsKansas2006).] The final set of columns compares episode-level datasets beginning with the original ICB project [@brecherInternationalCrisisBehavior2021; @brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020]; the Militarized Interstate Disputes dataset [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018], and the Correlates of War [@sarkeesResortWar181620072010]. With the exception of large-scale CAMEO dictionary-based systems, the existing state of the art quantitative datasets ignore the vast majority of the information content found in international relations.^[See @balaliCOfEEComprehensiveOntology2021 for a recent review of ontological depth and availability of Gold Standard example text.]

<!-- removedcites @braithwaiteMIDLOCIntroducingMilitarized2010; @braithwaiteCodebookMilitarizedInterstate2009; -->

@@

lit_review_alt <- lit_review_pull %>% 
                  mutate(Literature = paste0("",Literature %>% stringr::str_replace("; {0,}$","") ,"")) %>%
                  mutate(Literature = paste0("",Literature %>% stringr::str_replace_all(";",",") ,"")) %>%
                  #mutate(Literature = paste0("  [",Literature %>% stringr::str_replace("; {0,}$","") ,"]")) %>%
                  #mutate(Concept= paste0("**", Concept, "**", Literature)) %>%
                  mutate(Concept= paste0(Concept)) %>%
                  mutate(Literature = Literature %>% stringr::str_replace_all(stringr::fixed("NA"),"") ) %>%
                  mutate(Concept = Concept %>% stringr::str_replace(stringr::fixed(" ["),stringr::fixed(" - ["))) %>%
                  dplyr::select(-Codebook, -`Cameo Codes`, -`ICB Codes`, Literature, -`ICB Corpus`, -`Removed Literature`) # %>% 
                  # dplyr::rename(`Illustrative Literature` = Literature)
                  #dplyr::filter(is.na(group))

@@

We release several open-source products along with supporting code and documentation to further advance the study of IR, event extraction, and natural language processing. The first is the International Crisis Behavior Events (ICBe) dataset, an event-level aggregation of what took place during the crises identified by the ICB project. These data are appropriate for statistical analysis of hard questions about the sequencing of events (e.g. escalation and de-escalation of conflicts).^[Using ICBe data @gannonOneIfLand2022 finds that cross-domain crises are shorter and less violent than same-domain crises.] Second, we provide a coder-level disaggregation with multiple codings of each sentence by experts and undergrads that allows for the introduction of uncertainty and human interpretation of events. Further, we release a direct mapping from the codings to the source text at the sentence level as a new resource for natural language processing. Finally, we provide a companion website that incorporates detailed visualizations of all of the data introduced here at www.crisisevents.org.
